waRne - Putting cricket analytics in a spin

### 2020-10-03 paper_rob_working_good.Rmd
push
push

---
title: "waRne - Putting cricket analytics in a spin"
authors: Robert(1), James (2)
output:
  pdf_document:
    toc: no
  html_document:
    toc: yes
    toc_float: yes
    keep_md: yes
  word_document:
    toc: yes
bibliography: references.bib
---

@@

By making data more accessible more advanced metrics more accessible fans data journalism in sports has grown in recent years. For example in [@horowitz2017nflscrapr] has enabled EPA to enter popular discussion among fans
^[https://twitter.com/minakimes/status/1205591497497817089]
^[https://twitter.com/KevinColePFF/status/1277974109175951368].

@@

```{r, echo=FALSE, warning=FALSE,message=FALSE,out.width = "80%"}
p1<-cricjam_data_big %>% 
	dplyr::filter(batsman_id == 267192, 
								!is.na(pace_spin)) %>% 
	group_by(pace_spin, earned_runs) %>% 
	summarise(total = n()) %>% 
	group_by(pace_spin) %>% 
	mutate(totals = sum(total),
				 percent = total/totals, 
				 earned_runs = as.factor(earned_runs)) %>% 
	ggplot(aes(y = pace_spin, x = percent, color = earned_runs))+
	geom_point() +
	labs(color = "Runs", 
			 x = "% Balls Faced", 
			 y = "Bowling Style", 
			 title = "Steve Smith's Distribution of Runs Against Spin V Pace")
p1 +theme_minimal()

@@

```{r, echo=FALSE, message=FALSE,warning=FALSE,out.width = "80%"}

@@

Fans of cricket, instead of being mislead by [small samples](https://www.espncricinfo.com/story/_/id/30010174/kkr-vs-rr-ipl-2020-talking-points-why-teams-bowling-first?-did-andre-russell-promotion-work^[This article, has 6 matches being considered overwhelming evidence] might want to instead do a statistical test or visualise if batting first has a difference in terms of winning percentages and is this consistent across leagues.

```{r, echo=FALSE, warning=FALSE, message=FALSE,out.width = "80%"}
library(broom)
cricjam_data_big %>% 
	dplyr::filter(inning == 1, 
								batting_team_result != "super-over") %>% 
	dplyr::filter(tournament_type == "Twenty20", 
                tournament_name %in% c("Twenty20 Big Bash", "Indian Premier League", "Big Bash League", 
                                       "Champions League Twenty20", "Pepsi Indian Premier League")) %>% 
	mutate(t_name = if_else(tournament_name %in% c("Big Bash League", "Twenty20 Big Bash"), "Big Bash", "IPL")) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(year, batting_team_result, t_name) %>% 
	dplyr::select(batting_team_result, match_id, year, t_name) %>% 
	distinct() %>% 
	summarise(total = n()) %>% 
	pivot_wider(names_from = batting_team_result, values_from = total) %>% 
	ungroup() %>% 
	mutate(n =win+lose, 
				 p = 0.5, 
				 alternative = "two.sided", 
				 conf.level = 0.95, 
				 x = win) %>% 
	mutate(result = pmap(list(x, n,p), binom.test)) %>% 
  # Extract results from test into dataframe using broom::tidy
  mutate(result = map(result, tidy)) %>%
  mutate(result = map(result, ~select(.x, p.value, conf.low, conf.high))) %>% 
  unnest(cols = c(result)) %>%
  # plot the annual confidence intervals
  ggplot(aes(factor(year), p.value)) + 
    geom_errorbar(aes(ymin=conf.low, ymax=conf.high, color = (p.value > 0.05)), width=.1) +
  theme_minimal() +
  labs(y = "P(Batting is fair)", x = "Year", color = "p > 0.05") +
  ggtitle("Annual 1st Inning Batting Fairness")+ 
  theme(axis.text.x=element_text(angle=-90))+
	facet_wrap(~t_name)

@@

The yorker is considered one of most difficult deliveries to hit for batsman for a few reasons. Yorkers are usually bowled quicker than a bowlers average pace and aimed at the batsmans feet. This makes it difficult for the batsman to get ''under'' the ball and hit it over the fence for six. By pulling the delivery type from the play description on espncricinfo, every time a yorker was bowled we have the players who bowled and batted. As yorkers are typically bowled during the closing stages of matches, we can look at how good closers such as [Dhoni](https://www.espncricinfo.com/india/content/player/28081.html) bat against them.
```{r, echo=FALSE, message=FALSE, warning=FALSE,out.width = "80%"}
middle_batsman_of_interest <- c(
  230559, #pollard
  32498, #yusef pathan
  321777, #david miller
  44936, #AB De Villiers
  276298, #andre russell
  8180, #shane watson
  34102, #rohit sharma
  28081, #Dhoni
  325026 #maxwell
  )

@@

Cricket is an obvious sort of sport, to win in the end you have to get one run more than your opposition. A commonly used statistic during the broadcast is the required run rate, for example if you need 120 to win off 20 overs your RRR is $6$ runs an over. What follows is then obvious, the chasing team wins if they are able to lower the RRR to 0.

@@

```{r,message=FALSE, echo=FALSE, warning=FALSE,out.width = "80%"}
# graph of powerplay

@@

```{r,message=FALSE, echo=FALSE, warning=FALSE,out.width = "80%"}
# middle overs (enter in the middle)

@@

It seems as though in the middle he is letting the game get out of hand. How can one of the worlds best closers let the run rate go up so consistently in the middle overs?

@@

```{r,message=FALSE, echo=FALSE, warning=FALSE,out.width = "80%"}

@@

```{r, message=FALSE,echo=FALSE, warning=FALSE,out.width = "80%"}

@@

Finally we looked at the performance in the end of the game, overs 17-20. The required run rate is a poor metric to use here when the resources (overs and wickets) are low, so we restrict this to game situations where the [DLS](https://en.wikipedia.org/wiki/Duckworth%E2%80%93Lewis%E2%80%93Stern_method)^[The Duckworth Lewis tables are not publically accessible, through this example we hope someone somewhere will upload the tables for T20, for this example we had to use the ODI ones] resources are greater than 10. Here we can see that Dhoni is quickly lowering the required run rate, consistently to the end of the match. It appears that this is his strength, being able to at the last minute quickly lower the run near the end of the game in T20

```{r,message=FALSE,echo=FALSE, warning=FALSE,out.width = "80%"}

>>>>

---
title: "waRne - Putting cricket analytics in a spin"
authors: "unknown 1", "unknown 2"
output:
  pdf_document:
    toc: no
  html_document:
    toc: yes
    toc_float: yes
    keep_md: yes
  word_document:
    toc: yes
bibliography: references.bib
---

@@

By making data available more advanced metrics have become more accessible to  fans through quality data journalists. For example in [@horowitz2017nflscrapr] has enabled EPA to enter popular discussion among fans
^[https://twitter.com/minakimes/status/1205591497497817089]
^[https://twitter.com/KevinColePFF/status/1277974109175951368].

@@

```{r, echo=FALSE, warning=FALSE,message=FALSE,out.width = "80%",fig.align='center'}
p1<-cricjam_data_big %>% 
	dplyr::filter(batsman_id == 267192, 
								!is.na(pace_spin)) %>% 
	group_by(pace_spin, earned_runs) %>% 
	summarise(total = n()) %>% 
	group_by(pace_spin) %>% 
	mutate(totals = sum(total),
				 percent = total/totals, 
				 earned_runs = as.factor(earned_runs)) %>% 
	ggplot(aes(y = pace_spin, x = percent, color = earned_runs))+
	geom_point() +
	labs(color = "Runs", 
			 x = "% Balls Faced", 
			 y = "Bowling Style", 
			 title = "Steve Smith's Distribution of Runs Against Spin V Pace")
p1 +theme_minimal()

@@

```{r, echo=FALSE, message=FALSE,warning=FALSE,out.width = "80%",fig.align='center'}

@@

Fans of cricket, instead of being mislead by [small samples](https://www.espncricinfo.com/story/_/id/30010174/kkr-vs-rr-ipl-2020-talking-points-why-teams-bowling-first?-did-andre-russell-promotion-work)^[This article, has 6 matches being considered overwhelming evidence] might want to instead do a statistical test or visualise if batting first has a difference in terms of winning percentages and is this consistent across leagues.

```{r, echo=FALSE, warning=FALSE, message=FALSE,out.width = "80%",fig.align='center'}
library(broom)
cricjam_data_big %>% 
	dplyr::filter(inning == 1, 
								batting_team_result != "super-over") %>% 
	dplyr::filter(tournament_type == "Twenty20", 
                tournament_name %in% c("Twenty20 Big Bash", "Indian Premier League", "Big Bash League", 
                                       "Champions League Twenty20", "Pepsi Indian Premier League")) %>% 
	mutate(t_name = if_else(tournament_name %in% c("Big Bash League", "Twenty20 Big Bash"), "Big Bash", "IPL")) %>% 
	mutate(year = if_else(nchar(year)==4, year, as.numeric(str_sub(as.character(year), 1, 4)))) %>% 
	group_by(year, batting_team_result, t_name) %>% 
	dplyr::select(batting_team_result, match_id, year, t_name) %>% 
	distinct() %>% 
	summarise(total = n()) %>% 
	pivot_wider(names_from = batting_team_result, values_from = total) %>% 
	ungroup() %>% 
	mutate(n =win+lose, 
				 p = 0.5, 
				 alternative = "two.sided", 
				 conf.level = 0.95, 
				 x = win) %>% 
	mutate(result = pmap(list(x, n,p), binom.test)) %>% 
  # Extract results from test into dataframe using broom::tidy
  mutate(result = map(result, tidy)) %>%
  mutate(result = map(result, ~select(.x, p.value, conf.low, conf.high))) %>% 
  unnest(cols = c(result)) %>%
  # plot the annual confidence intervals
  ggplot(aes(factor(year), p.value)) + 
    geom_errorbar(aes(ymin=conf.low, ymax=conf.high, color = (p.value > 0.05)), width=.1) +
  theme_minimal() +
  labs(y = "P(Batting is fair)", x = "Year", color = "p < 0.05") +
  ggtitle("Annual 1st Inning Batting Fairness")+ 
  theme(axis.text.x=element_text(angle=-90))+
	facet_wrap(~t_name)

@@

The yorker is considered one of most difficult deliveries to hit for batsman for a few reasons. Yorkers are usually bowled quicker than a bowlers average pace and aimed at the batsmans feet. This makes it difficult for the batsman to get ''under'' the ball and hit it over the fence for six. By pulling the delivery type from the play description on espncricinfo, every time a yorker was bowled we have the players who bowled and batted. As yorkers are typically bowled during the closing stages of matches, we can look at how good closers such as [Dhoni](https://www.espncricinfo.com/india/content/player/28081.html) bat against them.

```{r, echo=FALSE, message=FALSE, warning=FALSE,out.width = "80%",fig.align='center'}
middle_batsman_of_interest <- c(
  230559, #pollard
  32498, #yusef pathan
  321777, #david miller
  44936, #AB De Villiers
  276298, #andre russell
  8180, #shane watson
  34102, #rohit sharma
  28081, #Dhoni
  325026 #maxwell
  )

@@

Cricket is an obvious sort of sport, to win in the end you have to get one run more than your opposition. A commonly used statistic during the broadcast is the required run rate (RRR), for example if you need 120 to win off 20 overs your RRR is $6$ runs an over. What follows is then obvious, the chasing team wins if they are able to lower the RRR to 0.

@@

```{r,message=FALSE, echo=FALSE, warning=FALSE,out.width = "80%",fig.align='center'}
# graph of powerplay

@@

```{r,message=FALSE, echo=FALSE, warning=FALSE,out.width = "80%",fig.align='center'}
# middle overs (enter in the middle)

@@

It seems as though in the middle he is letting the game get out of hand. How can one of the worlds best closers let the required run rate go up so consistently in the middle overs?

@@

```{r,message=FALSE, echo=FALSE, warning=FALSE,out.width = "80%",fig.align='center'}

@@

```{r, message=FALSE,echo=FALSE, warning=FALSE,out.width = "80%",fig.align='center'}

@@

Finally we looked at the performance in the end of the game, overs 17-20. The required run rate is a poor metric to use here when the resources (overs and wickets) are low, so we restrict this to game situations where the [DLS](https://en.wikipedia.org/wiki/Duckworth%E2%80%93Lewis%E2%80%93Stern_method)^[The Duckworth Lewis tables are not publically accessible, through this example we hope someone somewhere will upload the tables for T20, for this example we had to use the ODI ones] resources are greater than 10. Here we can see that Dhoni is quickly lowering the required run rate, consistently to the end of the match. It appears that this is his strength, being able to at the last minute quickly lower the required run rate near the end of the game in T20

```{r,message=FALSE,echo=FALSE, warning=FALSE,out.width = "80%",fig.align='center'}
